cubble An R Package for Structuring Spatio-temporal Data

### 2022-04-21 2-0-cube.Rmd
draft on cube

Multivariate spatio-temporal data can be conceptualised using a cubical data model with reference to variables, time and space. This naturally motivates the usage of array to represent and store spatio-temporal data, as evident by satellite imageries, large climate models, among others. The work by @lu_multidimensional_2018 has discussed the suitability of this representation through array algebra and practical spatio-temporal tasks.

The cubical conceptual framework provides a generalisation to the operation can be done on the data for visualisation. A paper by @bach_review_2014 has reviewed the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article "does not need to involve spatial data", but referring to "an abstract 2D substrate that is used to visualize data at a specific time". Despite its main focus is on temporal data, the mindset of abstracting out data representation to construct visualisation is applicable to our spatio-temporal data manipulation and visualisation.

>>>>

Spatio-temporal data can be conceptualised using a cubical data model with three axis: variable, time, and space. This abstraction can be a useful tool to map operations on spatio-temporal data to operations on the cube. @lu_multidimensional_2018 has shown an example of how array operations can be mapped onto the cube including select, scale, reduce, rearrange, and compute. The cubical abstract can also be used to generalise visualisation, where @bach_review_2014 has reviewed the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article "does not need to involve spatial data", but referring to "an abstract 2D substrate that is used to visualize data at a specific time". Despite its main focus is on temporal data, the mindset of abstracting out data representation to construct visualisation is applicable to our spatio-temporal data manipulation and visualisation.

While a cubical conceptual data model is neat to conceptually view the spatio-temporal data, data wrangling is not convenient in a cubical representation or arrays. There are a few reasons for this: 1) Some variables only have the spatial dimension and an array representation will need to handle the duplication issue mentioned above. 2) While arrays are efficient for computation on numerical variables, spatio-temporal data contains more than just numerical variables. Wrangling on character strings and specific datetime classes are not easy with arrays. 3) The 2D operation of creating a new variable, i.e. `mutate()`, is not clear defined in an array. There are three different ways to add a 1D vector to a cubical array and this doesn't precisely define how the new variable should be added.

### 2022-04-21 arxiv/cubble-arxiv.tex
draft on cube

Another issue is how the concept of tidy data (\protect\hyperlink{ref-tidydata}{Wickham 2014}) should be applied to spatio-temporal data. The tidy data concept advocates the data to be shaped into 1) one row per observation; 2) one column per variable; and 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e.~\pkg{dplyr} and \pkg{ggplot2} for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over long time period. This poses the question on how to arrange spatial and temporal variables that makes analysing spatio-temporal data easier.

@@

Spatio-temporal data can be conceptualised using a cubical data model with reference to variables, time and space. This naturally motivates the usage of array to represent and store spatio-temporal data, as evident by satellite imageries, large climate models, among others. The work by Lu, Appel, and Pebesma (\protect\hyperlink{ref-lu_multidimensional_2018}{2018}) has discussed the suitability of this representation through array algebra and practical spatio-temporal tasks. \textcolor{magenta}{SZ: is this too direct?}

The cubical conceptual framework provides a generalisation to the operation can be done on the data for visualisation. A paper by Bach et al. (\protect\hyperlink{ref-bach_review_2014}{2014}) has reviewed the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article ``does not need to involve spatial data'', but referring to ``an abstract 2D substrate that is used to visualize data at a specific time''. Despite its main focus is on temporal data, the mindset of abstracting out data representation to construct visualisation is applicable to our spatio-temporal data manipulation and visualisation.

@@

>>>>

Another issue is how the concept of tidy data (\protect\hyperlink{ref-tidydata}{Wickham 2014}) should be applied to spatio-temporal data. The tidy data concept advocates the data to be shaped into 1) one row per observation; 2) one column per variable; and 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e.~\pkg{dplyr} (\protect\hyperlink{ref-dplyr}{Wickham et al. 2022}) and \pkg{ggplot2} (\protect\hyperlink{ref-ggplot2}{Wickham 2016}) for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over long time period. This poses the question on how to arrange spatial and temporal variables that makes analysing spatio-temporal data easier.

@@

Spatio-temporal data can be conceptualised using a cubical data model with three axis: variable, time, and space. This abstraction can be a useful tool to map operations on spatio-temporal data to operations on the cube. Lu, Appel, and Pebesma (\protect\hyperlink{ref-lu_multidimensional_2018}{2018}) has shown an example of how array operations can be mapped onto the cube including select, scale, reduce, rearrange, and compute. The cubical abstract can also be used to generalise visualisation, where Bach et al. (\protect\hyperlink{ref-bach_review_2014}{2014}) has reviewed the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article ``does not need to involve spatial data'', but referring to ``an abstract 2D substrate that is used to visualize data at a specific time''. Despite its main focus is on temporal data, the mindset of abstracting out data representation to construct visualisation is applicable to our spatio-temporal data manipulation and visualisation.

While a cubical conceptual data model is neat to conceptually view the spatio-temporal data, data wrangling is not convenient in a cubical representation or arrays. There are a few reasons for this: 1) Some variables only have the spatial dimension and an array representation will need to handle the duplication issue mentioned above. 2) While arrays are efficient for computation on numerical variables, spatio-temporal data contains more than just numerical variables. Wrangling on character strings and specific datetime classes are not easy with arrays. 3) The 2D operation of creating a new variable, i.e.~\texttt{mutate()}, is not clear defined in an array. There are three different ways to add a 1D vector to a cubical array and this doesn't precisely define how the new variable should be added.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
---------. 2016. \emph{Ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.

\leavevmode\vadjust pre{\hypertarget{ref-dplyr}{}}%
Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2022. \emph{Dplyr: A Grammar of Data Manipulation}. \url{https://CRAN.R-project.org/package=dplyr}.
